WASP-4 is Accelerating Towards the Earth

### 2020-01-29 paper/ms.tex
update paper
derived a scaling relation

\begin{abstract}
  Space and ground-based transit observations of the hot Jupiter
  WASP-4b have recently shown that its orbital period is decreasing.
  The period change could be caused by tidal orbital decay, orbital
  precession, or light-travel time effects.  In this work we present
  new radial velocity measurements of WASP-4, acquired with
  Keck-HIRES.  The data show that the system is accelerating towards
  the Earth at $\dot{\gamma} = -0.0422^{+0.0028}_{-0.0027}\ {\rm
  m}\,{\rm s}^{-1}\,{\rm yr}^{-1}$.  This implied period decrease
  probably explains WASP-4's changing orbital period as a light-travel
  time effect.  By combining the radial velocities with new speckle
  imaging limits, we find that system's acceleration is likely caused
  by a 10-200$\,M_{\rm Jup}$ companion with semi-major axis between
  10-100$\,$AU.  Based on statistics from
  \citet{knutson_friends_2014}, about 1 in 6 hot Jupiters are expected
  to show period decreases commensurate with WASP-4 due to
  acceleration by outer companions.  Continued long-term radial
  velocity monitoring of hot Jupiters is therefore essential to
  distinguish tidal orbital decay from line-of-sight accelerations.
\end{abstract}

@@

We considered two models for the observed transit times.  The first
model assumes a constant orbital period on a circular orbit:
\begin{align}
  t_{\rm tra}(E) &= t_0 + PE,
\end{align}
where $E$ is the epoch number and $t_0$ is a reference epoch.  The
second model assumes that the period changes at a steady rate:
\begin{align}
  t_{\rm tra}(E) &=
    t_0 + PE +
    \frac{1}{2} \frac{{\rm d}P}{{\rm d}E} E^2.
\end{align}
The free parameters are the reference epoch $t_0$, the period at the
reference epoch $P$, and the period derivative, ${\rm d}P/{\rm d}t =
(1/P) {\rm d}P/{\rm d}E$.  We defined the epoch numbers such that
$E=0$ is near the weighted average of the observed times.  This helps
to reduce the covariance between $t_0$ and $P$.  A third possible
model that we did not consider for reasons that will become apparent
is a precessing, eccentric orbit \citep[{\it
e.g.},][]{gimenez_revision_1995,patra_2017}.

@@

For hot Jupiters that have been monitored for over 10 years, secular
changes in the orbital period be measured to a precision of $\lesssim
10~{\rm ms}\,{\rm yr}^{-1}$ ({\it e.g.}, K.~Patra et al{.} 2020,
submitted).

%FIXME
As observations from TESS continue...
As the Transiting Exoplanet

Continued long-term radial
  velocity monitoring of hot Jupiters is therefore essential to
  distinguish tidal orbital decay from line-of-sight accelerations.

@@

Over half of hot Jupiters have outer companions with masses larger
than Jupiter outside of 1$\,$AU. The accelerations and period changes
induced by these outer companions will become an increasingly large
nuisance in the hunt for orbital decay as the observational baselines
get longer.  This issue can be alleviated through long-term radial
velocity monitoring of hot Jupiters.

%TODO: what's the scaling? sigma_P ~ 1/baseline? baseline^2?

>>>>

\begin{abstract}
  Space and ground-based transit observations of the hot Jupiter
  WASP-4b have recently indicated that its orbital period is decreasing.
  The period change could be caused by tidal orbital decay, orbital
  precession, or light-travel time effects.  In this work we present
  new radial velocity measurements of WASP-4, acquired with
  Keck-HIRES.  The data show that the system is accelerating towards
  the Earth at $\dot{\gamma} = -0.0422^{+0.0028}_{-0.0027}\ {\rm
  m}\,{\rm s}^{-1}\,{\rm yr}^{-1}$.  This implied period decrease
  probably explains WASP-4's changing orbital period as a light-travel
  time effect.  By combining the radial velocities with new limits
  from speckle imaging, we find that system's acceleration is likely
  caused by a 10-200$\,M_{\rm Jup}$ companion with semi-major axis
  between 10-100$\,$AU.  Based on statistics from
  \citet{knutson_friends_2014}, about 1 in 6 hot Jupiters are expected
  to show period decreases commensurate with WASP-4 due to
  acceleration by outer companions.  These period decreases will be
  increasingly measureable in coming years, because the precision of
  period derivative measurements scales as the fourth power of the
  observing baseline.  Continued long-term radial velocity monitoring
  of hot Jupiters is therefore essential to distinguish tidal orbital
  decay from line-of-sight accelerations.
\end{abstract}

@@

We considered two models for the observed transit times.  The first
model assumes a constant orbital period on a circular orbit:
\begin{align}
  t_{\rm tra}(E) &= t_0 + PE,
\end{align}
where $E$ is the epoch number and $t_0$ is a reference epoch.  The
second model assumes that the period changes at a steady rate:
\begin{align}
  t_{\rm tra}(E) &=
    t_0 + PE +
    \frac{1}{2} \frac{{\rm d}P}{{\rm d}E} E^2.
  \label{eq:quadratic_model}
\end{align}
The free parameters are the reference epoch $t_0$, the period at the
reference epoch $P$, and the period derivative, ${\rm d}P/{\rm d}t =
(1/P) {\rm d}P/{\rm d}E$.  We defined the epoch numbers such that
$E=0$ is near the weighted average of the observed times.  This helps
to reduce the covariance between $t_0$ and $P$.  A third possible
model that we did not consider for reasons that will become apparent
is a precessing, eccentric orbit \citep[{\it
e.g.},][]{gimenez_revision_1995,patra_2017}.

@@

\subsection{At what rate is the measurement precision of d$P$/d$t$
increasing?}
\label{sec:fisher}

For hot Jupiters that have been monitored for over baselines $\tau$ of
$\approx$10 years, secular changes in their orbital periods are
currently being constrained to a precision of $\lesssim 10~{\rm
ms}\,{\rm yr}^{-1}$ ({\it e.g.}, K.~Patra et al{.} 2020, submitted).
This is roughly commensurate with level of signal many outer
companions are expected to induce from radial velocity accelerations
(Figure~\ref{fig:pdot_pop}).

It is therefore naturally of interest to ask at what point in time
additional detections of the light-travel time effect can become
common-place among hot Jupiters.  This question is the same as
asking: at what rate does the uncertainty in the quadratic term of
Equation~\label{eq:quadratic_model} scale with the observing baseline?
We take a Fisher analysis approach to the problem, following the notes
by \citet{gould_chi2_2003}.
Rewriting Equation~\ref{eq:quadratic_model} in simpler terms,
\begin{align}
  t_{\rm tra} = a_0 + a_1\cdot E + a_2\cdot E^2,
\end{align}
where $a_0=t_0$, $a_1=P$, $a_2=0.5\cdot {\rm d}P/{\rm d}E$.
One can then show that if $N$ transit time measurements are taken at
uniform times across a baseline of duration $\tau$ with constant
precision $\sigma_{\rm tra}$, then the uncertainty of the quadratic
term is given by
\begin{align}
  \sigma_{a_2} =
  \left(\frac{25{,}920}{N}\right)^{1/2} \frac{\sigma_{\rm tra}}{\tau^4}
\end{align}
This result implies that if a decade passes, doubling the observing
baseline, we gain a factor of 16 in precision on d$P$/d$t$.  Assuming
regular observations continue from ground-based telescopes (CITE) and
space-based observatories (CITE TESS, CITE CHEOPS), this will enable
measurements of period derivatives precise to less than $1~{\rm
ms}\,{\rm yr}^{-1}$ before 2030.
%FIXME the units above make no sense????

@@

Over half of hot Jupiters have outer companions with masses larger
than Jupiter that orbit beyond 1$\,$AU
\citep{knutson_friends_2014,bryan_statistics_2016}. The accelerations
and period changes induced by these outer companions will become an
increasingly large nuisance in the hunt for orbital decay as the
observational baselines get longer.  In particular, the precision with
which the period derivative can be measured scales as the fourth power
of the baseline duration (\S~\ref{sec:fisher}), and so within a decade
many more hot Jupiters should show orbital period changes due to
accelerations induced by their outer companions.  This effect can be
distinguished from tidal orbital decay through long-term radial
velocity monitoring of hot Jupiters.
